Analysis and Figures for Gallego et al, 2020

### 2020-09-06 Manuscript/What.lies.ahead.reviewed.tex
intermediate changes for Moncho's review. still working on Results.

These communities were distinguished by changes in the relative abundances of a wide variety of taxa. In the Hood Canal for example, the colder water and higher-pH community (community 1 in Fig. 2C) showed higher eDNA indexes of diatoms \textit{Thalassiosira nordenskioeldii}, \textit{Nitzschia cf. promare}, and \textit{Ditylum brightwellii} -- among other species -- relative to a higher-temperature community (community 3 in Fig. 2C). Communities 2 and 3 had relatively higher values for taxa from a mix of groups including \textit{Emiliana huxleyi}, \textit{Minutocellus polymorphus}, \textit{Micromonas pusilla}, and \textit{Bathycoccus prasinos}. Community 3 (Fig. 2A,B) and similar planktonic communities occupy the spectrum of environmental conditions most likely to be encountered in near-future climate scenarios as temperature rises and pH falls. For example, by 2095 we expect summer in the Hood Canal to be T $>$ 12\degree C and pH $<$ {7.9} -- where community 3 is the most likely community -- to occur more than {four-fifths} of the time (83\%}). {\color{blue} I couldn't find where you were doing this calculation, so I couldn't do the fractions of time we expect the other conditions. It would be good to say that we go from condition 1 being XX\% of the time in 2017 to YY\% of the time in 2095.}

@@

Given such heterogeneity in projections, gains and losses tend to balance one another out when looking at overall richness variation; {\color{red} we find no change in median richness for the year 2095 relative to the present in the Hood Canal (overall taxon richness by year, averaged over all environmental conditions, Wilcoxon p $=$ 0.5); while higher diversity is expected in the San Juan Island in 2095 (increase in median species richness of 2.1, Wilcoxon p $=$ 0). Diatoms, however, show small but significant declines in richness in the Hood Canal (0.5 species, Wilcoxon p $< 10^{-16}$), while the changes on the San Juan Island are undetectable (p $=$ 0.8).  Dinoflagellates see their richness increase in both regions with the future scenario(Wilcoxon p $< 10^{-50}$) }.  {\color{blue} I re-did some of these in the Rmd file (starting at line 1129, doing it line-by-line to convince myself what was going on. See line 1157 in particular: for diatoms, there is no change between years, but yet the wilcox p value is very low. I can't figure out what's happening there.}

@@

{\color{blue} The paragraph below is important, and links to the question I had above: where are we doing this calculation in the code? and is the below paragraph still true? Plus, a reviewer wanted a citation of the Coscinodiscus statement.}

The bulk of our projected community changes result from now-rare conditions occurring more frequently in the future. For example, in summer in the Hood Canal at present, we expect surface waters to have pH $<$ 7.9 and T $>$ 19$\degree$C only 1\% of the time. In 2095, we expect these conditions 18 times more frequently (i.e., 18\% of the time). At these values of T and pH, our model predicts the harmful \textit{Alexandrium} sp. to occur more often than not (mean frequency of occurrence $=$ 0.7). By contrast, the large centric diatom \textit{Coscinodiscus} -- a key source of carbon for zooplankton and small fishes -- occurs only one-third of the time under these same conditions (mean frequency $=$ 0.34).

>>>>

These communities were distinguished by changes in the relative abundances of a wide variety of taxa. In the Hood Canal for example, the colder water and higher-pH community (community 1 in Fig. 2C) showed higher eDNA indexes of diatoms \textit{Thalassiosira nordenskioeldii}, \textit{Nitzschia cf. promare}, and \textit{Ditylum brightwellii} -- among other species -- relative to a higher-temperature community (community 3 in Fig. 2C). Communities 2 and 3 had relatively higher values for taxa from a mix of groups including \textit{Emiliana huxleyi}, \textit{Minutocellus polymorphus}, \textit{Micromonas pusilla}, and \textit{Bathycoccus prasinos}. Community 3 (Fig. 2A,B) and similar planktonic communities occupy the spectrum of environmental conditions most likely to be encountered in near-future climate scenarios as temperature rises and pH falls. For example, we expect the Hood Canal to be T $>$ 12\degree C and pH $<$ {7.9} -- where community 3 is the most likely community -- more than four times as often in 2095 than in 2017 (4.9 \% vs 20.7\%). {\color{blue} I couldn't find where you were doing this calculation, so I made one up -- see lines 1134 and following in the Rmd. BUT ALSO : community 3 isn't sensitive to pH, so what are we even saying here?}

@@

Given such heterogeneity in projections, gains and losses tend to balance one another out when looking at overall richness variation; {\color{red} we find no change in median richness for the year 2095 relative to the present in the Hood Canal (overall taxon richness by year, averaged over all environmental conditions, Wilcoxon p $=$ 0.5); while higher diversity is expected in the San Juan Island in 2095 (increase in median species richness of 2.1, Wilcoxon p $< 10^{-16}$). Diatoms, however, show small but significant declines in richness in the Hood Canal (0.5 species, Wilcoxon p $< 10^{-16}$), while the changes on the San Juan Island are undetectable (p $=$ 0.8).  Dinoflagellates see their richness increase in both regions with the future scenario(Wilcoxon p $< 10^{-16}$) }.  {\color{blue} I re-did some of these in the Rmd file (starting at line 1129, doing it line-by-line to convince myself what was going on. See line 1179 in particular: for diatoms, there is no change between years, but yet the wilcox p value is very low. I can't figure out what's happening there.}

@@

{\color{blue} The paragraph below is important, and links to the question I had above: where are we doing this calculation in the code? I went through line by line; double check calculations -- see lines 1207 and following in the Rmd. Plus, a reviewer wanted a citation of the Coscinodiscus statement, so I added some.}

{\color{blue}The bulk of our projected community changes result from now-rare conditions occurring more frequently in the future. For example, in in the Hood Canal at present, we expect surface waters to have pH $<$ 7.9 and T $>$ 19$\degree$C only 1\% of the time. In 2095, we expect these conditions 6 times more frequently (i.e., 6\% of the time). At these values of T and pH, our model predicts the harmful \textit{Alexandrium} sp. to occur more often than not (mean frequency of occurrence $=$ 0.83). By contrast, the large centric diatom \textit{Coscinodiscus} -- a potentially key source of carbon for zooplankton and small fishes \cite{pasquaud2010determination, zamon2002tidal} with effects on dissolved oxygen and other water-column characteristics \cite{manabe1991bloom} -- occurs only one-third of the time under these same conditions (mean frequency $=$ 0.35).}

### 2020-09-06 Scripts/Manuscript.Main.Analysis.Rmd
intermediate changes for Moncho's review. still working on Results.

@@

#while higher diversity is expected in the San Juan Island in 2095 (increase in median species richness of 2.1, Wilcoxon p $=$ 0).

@@

#Dinoflagellates see their richness increase in both regions with the future scenario(Wilcoxon p $< 10^{-50}$)

>>>>

#For example, we expect the Hood Canal to be T $>$ 12\degree C and pH $<$ {7.9} -- where community 3 is the most likely community -- more than four times as often in 2095 than in 2017 (4.9 \% vs 20.7\%) )

stds_to_value$toValue[[3]]
stds_to_value$toValue[[4]]

richness.by.phylum.by.point %>% 
  ungroup() %>% 
  dplyr::select(Area, Temperature, pH, Year) %>% 
  distinct() %>% 
  filter(Area == "Hood Canal") %>% 
  mutate(Temperature = 14.995 + 4.299*Temperature,
         pH = 8.2157 + 0.1813*pH) %>% 
  group_by(Year) %>% 
  add_count() %>% 
  summarise(condition = sum(Temperature > 12 & pH < 7.9),
            totalcount = n()) %>% 
  mutate(percent_time = condition / totalcount)

@@

#while higher diversity is expected in the San Juan Island in 2095 (increase in median species richness of 2.1, Wilcoxon p $<$ 10 ^{-16}).

@@

#Dinoflagellates see their richness increase in both regions with the future scenario(Wilcoxon p $< 10^{-16}$)

wilcox.test(x = richness.by.phylum.by.point %>%
                  filter(Year == 2017, Area == "Hood Canal", phylum == "Dinophyceae") %>% 
                    pull(richness),
            y = richness.by.phylum.by.point %>%
                  filter(Year == 2095, Area == "Hood Canal", phylum == "Dinophyceae") %>% 
                    pull(richness))

wilcox.test(x = richness.by.phylum.by.point %>%
                  filter(Year == 2017, Area == "San Juan Island", phylum == "Dinophyceae") %>% 
                    pull(richness),
            y = richness.by.phylum.by.point %>%
                  filter(Year == 2095, Area == "San Juan Island", phylum == "Dinophyceae") %>% 
                    pull(richness))

#For example, in in the Hood Canal at present, we expect surface waters to have pH $<$ 7.9 and T $>$ 19$\degree$C only 1\% of the time. In 2095, we expect these conditions 6 times more frequently (i.e., 6\% of the time).

stds_to_value$toValue[[3]]
stds_to_value$toValue[[4]]

richness.by.phylum.by.point %>% 
  ungroup() %>% 
  dplyr::select(Area, Temperature, pH, Year) %>% 
  distinct() %>% 
  filter(Area == "Hood Canal") %>% 
  mutate(Temperature = 14.995 + 4.299*Temperature,
         pH = 8.2157 + 0.1813*pH) %>% 
  group_by(Year) %>% 
  add_count() %>% 
  summarise(condition = sum(Temperature > 19 & pH < 7.9),
            totalcount = n()) %>% 
  mutate(percent_time = condition / totalcount)

#At these values of T and pH, our model predicts the harmful \textit{Alexandrium} sp. to occur more often than not (mean frequency of occurrence $=$ 0.833)

expanded.future %>% 
  filter(Area == "Hood Canal", Year == 2095, TemperatureCelsius > 19, pH_units < 7.90, taxa == "Gonyaulacaceae|Alexandrium|NA") %>% 
  dplyr::select(present) %>% 
  summarise(mean(present))

#By contrast, the large centric diatom \textit{Coscinodiscus} -- a potentially key source of carbon for zooplankton and small fishes \cite{pasquaud2010determination, zamon2002tidal} with effects on dissolved oxygen and other water-column characteristics \cite{manabe1991bloom} -- occurs only one-third of the time under these same conditions (mean frequency $=$ 0.35)
expanded.future %>% 
  filter(Area == "Hood Canal", Year == 2095, TemperatureCelsius > 19, pH_units < 7.90, taxa == "Coscinodiscaceae|Coscinodiscus|Coscinodiscus wailesii") %>% 
  dplyr::select(present) %>% 
  summarise(mean(present))
